---
title: "Adaptive non-parametric learning"
author: "Chris Holmes, Simon Lyddon, Miguel Morin, and James Robinson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adaptive non-parametric learning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Bayesian learning is built on an assumption that the model space contains a true
reflection of the data generating mechanism. This assumption is problematic,
particularly in complex data environments. Here we present a Bayesian
nonparametric approach to learning that makes use of statistical models, but
does not assume that the model is true. Our approach admits a Monte Carlo sampling
scheme that can afford massive scalability on modern computer architectures. The
model-based aspect of learning is particularly attractive for regularizing
nonparametric inference when the sample size is small, and also for correcting
approximate approaches such as variational Bayes (VB).

Here, we demonstrate the approach on a VB classifiers.

## Setup

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
