---
title: "Adaptive non-parametric learning"
author: "Chris Holmes, Simon Lyddon, Miguel Morin, and James Robinson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adaptive non-parametric learning}
  %\VignetteEngine{knitr::rmarkdown_notangle}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=7,
  fig.height=3
)
```

# Introduction

This vignette shows the code generating figure 2 in Lyddon, Walker & Holmes,
2018, and also reuses material from that paper.

Bayesian learning is built on an assumption that the model space contains a true
reflection of the data generating mechanism. This assumption is problematic,
particularly in complex data environments. Here we present a Bayesian
nonparametric approach to learning that makes use of statistical models, but
does not assume that the model is true. Our approach admits a Monte Carlo sampling
scheme that can afford massive scalability on modern computer architectures. The
model-based aspect of learning is particularly attractive for regularizing
nonparametric inference when the sample size is small, and also for correcting
approximate approaches such as variational Bayes.

Here, we demonstrate the approach on a variational Bayes classifier.

We demonstrate this in practice through a variational Bayes logistic regression model fit to
the Statlog German Credit dataset, containing 1000 observations and 25
covariates (including intercept), from the [UCI ML
repository](http://archive.ics.uci.edu/ml/datasets/Statlog+%28German+Credit+Data%29)
and which ships with the package. The outcome is whether an individual has good
credit rating.

# Setup

## Dependencies

We require our package and the `rstan` package for variational Bayes. Note that,
for technical reasons (see [StackOverflow issue](https://stackoverflow.com/questions/56262828/function-passes-test-inside-the-package-and-fails-from-vignette-due-to-cpp-obje)), the `rstan` package needs to be loaded and attached, so
we use `library("rstan")`.

```{r, results = 'hide'}
requireNamespace("PosteriorBootstrap", quietly = TRUE)

requireNamespace("dplyr", quietly = TRUE)
requireNamespace("ggplot2", quietly = TRUE)
requireNamespace("tibble", quietly = TRUE)

library("rstan")
```

## Plotting functions

We define a `ggproto` object to compute the density of samples, and wrap it in a `ggplot2` object:

```{r}
#The first argument is required, either NULL or an arbitrary string.
stat_density_2d1_proto <- ggplot2::ggproto(NULL,
  ggplot2::Stat,
  required_aes = c("x", "y"),

  compute_group = function(data, scales, bins, n) {
    # Choose the bandwidth of Gaussian kernel estimators and increase it for
    # smoother densities in small sample sizes
    h <- c(MASS::bandwidth.nrd(data$x) * 1.5,
           MASS::bandwidth.nrd(data$y) * 1.5)

    # Estimate two-dimensional density
    dens <- MASS::kde2d(
      data$x, data$y, h = h, n = n,
      lims = c(scales$x$dimension(), scales$y$dimension())
    )

    # Store in data frame
    df <- data.frame(expand.grid(x = dens$x, y = dens$y), z = as.vector(dens$z))

    # Add a label of this density for ggplot2
    df$group <- data$group[1]

    # plot
    ggplot2::StatContour$compute_panel(df, scales, bins)
  }
)

# Wrap that ggproto in a ggplot2 object
stat_density_2d1 <- function(data = NULL,
                             geom = "density_2d",
                             position = "identity",
                             n = 100,
                             ...) {
    ggplot2::layer(
      data = data,
      stat = stat_density_2d1_proto,
      geom = geom,
      position = position,
      params = list(
        n = n,
        ...
      )
    )
}
```

We define a shorthand function that appends all samples to a dataframe and is ready for plotting.

```{r}
append_to_plot <- function(plot_df, sample, method,
                           concentration, x_index, y_index) {
  new_plot_df <- rbind(plot_df, tibble::tibble(x = sample[, x_index],
                                               y = sample[, y_index],
                                               Method = method,
                                               concentration = concentration))
  return(new_plot_df)
}
```

# Adaptive non-parametric learning

## Sampling

We assign an independent normal prior with standard deviation 10 to each covariate, and
generate 1000 posterior samples for each method: Bayesian logistic regression,
variational Bayes, and Adaptive Non-Parametric Learning.

**Note**: variational Bayes in RStan is still experimental. We have had situations in the past where the same code with the same versions produces very different results. [This thread on StackOverflow](https://stackoverflow.com/questions/57186920/rstan-gives-different-results-in-exact-and-variational-bayes-modes) mentions this problem. In this code, variational Bayes provides the samples for our method, so the results may vary in the future because of the results from variational Bayes in RStan. Whether variational Bayes gives the expected answers or not, our approach will always give an answer that is in-between the results from variational Bayes and Bayesian logistic regression.

We tune the settings for the sampling with the number of draws, setting the
seed, and getting the data:

```{r}
seed <- 123
prior_sd <- 10
n_bootstrap <- 1000
german <- PosteriorBootstrap::get_german_credit_dataset()
```

For Bayesian logistic regression, we draw samples using the `rstan` package and
the Bayesian logistic regression model that ships with this package:

```{r}
if ("Windows" != Sys.info()["sysname"]) {
  train_dat <- list(n = length(german$y), p = ncol(german$x), x = german$x, y = german$y, beta_sd = prior_sd)
  stan_file <- PosteriorBootstrap::get_stan_file()
  bayes_log_reg <- rstan::stan(stan_file, data = train_dat, seed = seed,
                        iter = n_bootstrap * 2, chains = 1)
  stan_bayes_sample <- rstan::extract(bayes_log_reg)$beta
}
```

For variational Bayes, we obtain a mean-field variational Bayes sample using
automatic differentiation variational inference using the `rstan` package and
the same logistic regression model as above. The
number of samples is `n_bootstrap`,  as these samples serve for the Adaptive
Non-Parametric Learning algorithm:

```{r}
if ("Windows" != Sys.info()["sysname"]) {
  stan_model <- rstan::stan_model(file = stan_file)
  stan_vb <- rstan::vb(object = stan_model, data = train_dat, seed = seed,
                  output_samples = n_bootstrap)
  stan_vb_sample <- rstan::extract(stan_vb)$beta
}
```

We now run the Adaptive Non-Parametric Learning algorithm, using the samples
from variational Bayes, for two different values of the concentration
hyper-parameter. Because our sampling method with `c = 1000` takes longer than the timeout allowed on CRAN, we use only `c = 500` for this vignette, but please see the [GitHub page](https://github.com/alan-turing-institute/PosteriorBootstrap) for the same figure as in the paper (with `c = 1`, `c = 1,000`, and `c = 20,000`).

```{r}
if ("Windows" != Sys.info()["sysname"]) {
  if (identical(Sys.getenv("NOT_CRAN", unset = "true"), "true")) {
    concentrations <- c(1, 500)
    anpl_samples <- list()
    for (concentration in concentrations) {
      anpl_sample <- PosteriorBootstrap::draw_logit_samples(x = german$x, y = german$y,
                                                            concentration = concentration,
                                                            n_bootstrap = n_bootstrap,
                                                            posterior_sample = stan_vb_sample,
                                                            threshold = 1e-8,
                                                            show_progress = TRUE)
      anpl_samples[[toString(concentration)]] <- anpl_sample
    }
  }
}
```

## Preparing the plot

We now prepare a dataframe with all the samples ready for plotting.

```{r}
if ("Windows" != Sys.info()["sysname"]) {
  if (identical(Sys.getenv("NOT_CRAN", unset = "true"), "true")) {
    # Initialise
    plot_df <- tibble::tibble()

    # Index of coefficients in the plot
    x_index <- 21
    y_index <- 22

    # Create a plot data frame with all the samples
    for (concentration in concentrations) {
      plot_df  <- append_to_plot(plot_df, sample = anpl_samples[[toString(concentration)]],
                                method = "PosteriorBootstrap-ANPL",
                                concentration = concentration,
                                x_index = x_index, y_index = y_index)
      plot_df  <- append_to_plot(plot_df, sample = stan_bayes_sample,
                                method = "Bayes-Stan",
                                concentration = concentration,
                                x_index = x_index, y_index = y_index)
      plot_df  <- append_to_plot(plot_df, sample = stan_vb_sample,
                                method = "VB-Stan",
                                concentration = concentration,
                                x_index = x_index, y_index = y_index)
    }
  }
}
```

## Plotting the results

And now we plot the result of the algorithm:

```{r, fig.show = "asis"}
if ("Windows" != Sys.info()["sysname"]) {
  if (identical(Sys.getenv("NOT_CRAN", unset = "true"), "true")) {
    ggplot2::ggplot(ggplot2::aes_string(x = "x", y = "y", colour = "Method"),
                    data = dplyr::filter(plot_df, plot_df$Method != "Bayes-Stan")) +
      stat_density_2d1(bins = 5) +
      ggplot2::geom_point(alpha = 0.1, size = 1,
                          data = dplyr::filter(plot_df,
                                              plot_df$Method == "Bayes-Stan")) +
      ggplot2::facet_wrap(~concentration, nrow = 1,
                          scales = "fixed",
                          labeller = ggplot2::label_bquote(c ~" = "~
                                                            .(concentration))
                          ) +
      ggplot2::theme_grey(base_size = 8) +
      ggplot2::xlab(bquote(beta[.(x_index)])) +
      ggplot2::ylab(bquote(beta[.(y_index)])) +
      ggplot2::theme(legend.position = "none",
                    plot.margin = ggplot2::margin(0, 10, 0, 0, "pt"))
  }
}
```

The Non-parametric update effectively corrects the variational Bayes
approximation for small values of the concentration `c`. As we increase the
concentration parameter, the samples drawn with our method move closer to those
from the variational Bayes approximation, showing that the concentration
parameter is an effective sample size tuning of the weight to give to the data
compared to the approximate model.

Of course, local variational methods can provide more accurate posterior
approximations to Bayesian logistic posteriors, though these too are
approximations, that our algorithm can correct.

# Parallelisation and performance

Our construction admits a trivially parallelisable sampler once the parametric
posterior samples have been generated (or if the concentration `c` equals
0). Note that, although some algorithms to generate the parametric posterior
samples may be parallelisable, above we use a Markov Chain-Monte Carlo algorithm
from RStan, which is not parallelisable. We emphasise that the part of our
algorithm that benefits from modern parallel computer architectures happens
after we have the posterior samples (or if `c=0`).

We now illustrate the parallelisation speedup. The calculation of the expected
speedup depends on the number of bootstrap samples and the number of
processors. It also depends on the system: it is larger on macOS than on Linux,
with some variation depending on the version of R.

Due to limitations in the R workflow,
[this vignette is limited to one or two
cores](https://stackoverflow.com/questions/41307178/error-processing-vignette-failed-with-diagnostics-4-simultaneous-processes-spa),
but see the [GitHub
page](https://github.com/alan-turing-institute/PosteriorBootstrap/blob/master/README.md)
for a plot up to 64 cores.

### Ahmdal's law

Fixing the number of samples corresponds to [Ahmdal's
law](https://en.wikipedia.org/wiki/Ahmdal's_Law), or the speedup in the task as
a function of the number of processors. The speedup `S_latency` of `N`
processors is defined as the duration of the task with one core divided by the
duration of the task with `N` processors.  We compute the duration for one or two cores and for
several values of bootstrap samples.

```{r}
if ("Windows" != Sys.info()["sysname"] ) {
  speedups  <- data.frame(stringsAsFactors = FALSE)
  max_cores <- 2
  n_bootstrap_array <- c(1e2, 2e2, 5e2)

  for (n_bootstrap in n_bootstrap_array) {
    one_core_duration <- NULL
    for (num_cores in seq(1, max_cores)) {

      start <- Sys.time()
      anpl_samples <- PosteriorBootstrap::draw_logit_samples(x = german$x, y = german$y,
                                                            concentration = 1,
                                                            n_bootstrap = n_bootstrap,
                                                            gamma_mean = rep(0, ncol(german$x)),
                                                            gamma_vcov = diag(1, ncol(german$x)),
                                                            threshold = 1e-8,
                                                            num_cores = num_cores)
      lap <- as.double((Sys.time() - start), units = "secs")
    if (1 == num_cores) {
      one_core_duration <- lap
      }
      speedups <- rbind(speedups, c(num_cores, n_bootstrap, one_core_duration / lap))
    }
    names(speedups) <- c("Num_cores", "N_bootstrap", "speedup")
  }

  # Convert n_bootstrap to strings for ggplot2 to arrange them into groups
  speedups$N_bootstrap <- paste0("N_", speedups$N_bootstrap)

  ggplot2::ggplot(data = speedups,
                  ggplot2::aes(x = Num_cores, y = speedup)) +
    ggplot2::geom_line(ggplot2::aes(colour = N_bootstrap))
}
```

We invert Ahmdal's law to compute the proportion of the execution time that is
parallelisable from the speedup as:

$$ p = \frac{\frac{1}{S_{latency}}} - 1}{\frac{1}{s} - 1} $$

where $S_{latency}$ is the theoretical speedup of the whole task in Ahmdal's law
and the observed speedup here, and $s$ is the speedup of the part of the task
that can be parallelised, and thus equal to the number of
processors. Calculating this value for the durations from 1 to 8 cores, we obtain
this plot, where the proportion of the code that can be parallelised is high:

```{r}
if ("Windows" != Sys.info()["sysname"] ) {
  library("gridExtra")

  # Remove single core speedup, where the proportion is not defined
  speedups <- speedups[1 != speedups$Num_cores, ]
  speedups$proportion <- (1 / speedups$speedup - 1) / (1 / speedups$Num_cores - 1)

  ggplot2::qplot(proportion, data = speedups, fill = N_bootstrap, binwidth = 0.005) +
    ggplot2::facet_wrap(facets = ~N_bootstrap)
}
```
